### 2017-10-24 paper/sexSelectionSpacing-ver1.tex
Introduce median duration and sex ration tables in text

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
This is because each sex-selective abortion substantially increases the 
spacing from last birth. 
For every abortion a woman has we should expect the duration from last
birth to increase by approximately a year.%
\footnote{
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

@@

The main downside of my approach is that direct interpretation of the estimated 
coefficients for this model is challenging because of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, 
no birth, rather than simply the hazard of an event.
Second, a positive coefficient does not necessarily imply that an increase in a
variable's value increases the probability of the associated event because the 
probability of another event may increase even more \citep{thomas96}.

@@

Spell duration is measured in quarters of a year, that is 3 month periods, hereafter
referred to as quarters.
The first spell begins at the month of marriage rather than 9 months after because many 
women report giving birth less than 9 months after they were married.
For women who began living with their husbands at too young an age to conceive the 
starting point should ideally be first ovulation, when she becomes ``at risk'' for a 
pregnancy, rather than month of marriage.
Unfortunately, information on age of menarche is only available in NFHS-1.
Instead, for women who began living with their husband before age 12, I set the 
the month they turned 12 years of age as staring point for the first spell.

@@

All spells continue until either a child is born or the spell is censored.
Censoring can happen for three reasons:
the survey takes place;
the woman is sterilized;
or the number of births observed becomes too sparse for the method to work.
The timing of censoring because of too few births vary slightly by spell but is generally 
21 quarters after the beginning of the spell.

@@

The area of residence is a dummy variable for the household living in
an urban area.%
\footnote{
NFHS uses four categories for area of residence: Large city, small city, town
and countryside.
To reach a sufficient sample size urban areas are merged into one group.
}
The cost of children is higher in urban areas than in rural and access to prenatal
sex determination is easier, and both are expected to be associated with
greater use of sex-selective abortions in urban than in rural areas.
Because of concerns about selective migration I use where the household was living
at the end of each spell.%
\footnote{
As online Appendix Table D.1 shows there is little difference across
education levels in migration patterns within each period and patterns across periods
are also relatively stable.
}

@@

The increase in urbanization and education is likely to exert downward pressure
on fertility and the high censoring rates for the later periods are evidence of this.
The average number of children born to women by the time they turn 35 illustrate how
strong the decline in fertility has been.%
\footnote{
Figure \ref{fig:fertility} shows the full results.
}
Women born in the early 1940s had on average close to 5 children when 
they reached 35, but women born in the early 1970s had only just over three children.
The low number of children is especially remarkable because it combines
all education levels and all areas of residence.
% Hence, fertility in cities is likely substantially lower.

@@

@@

Figure \ref{fig:results_spell2_high_urban} 
shows the predicted probabilities of having a boy and standard survival curves
for the three periods for urban women with 8 or more years of education.%
\footnote{
The corresponding graphs for rural women and for women with no education are
shown in Figures \ref{fig:results_spell2_high_rural}, \ref{fig:results_spell2_low_urban},
and \ref{fig:results_spell2_low_rural}
}
Two main results stand out.
First, women whose first child was a boy do not appear to use sex selection during
the second spell.
The predicted percent boys is consistently close to the natural level in all three
panels (\ref{fig:results_spell2_high_urban}(d)-(f)).
Second, as prenatal sex determination became available there is more and more
evidence of sex selection affecting the sex ratio for women with a girl as their
first child.
In panel (a), which covers the period before sex selection became generally
available, there is no significant evidence of a sex ratio biased towards males.
Panel (b) shows a more male biased sex ratio but only a few of quarters show
a statistically significant difference from the natural rate and the shallow 
slope of the survival curve indicates that relatively few births took place
during these quarters.
Panel (c), however, shows clear evidence of the effect of access to prenatal
sex determination with a male biased sex ratio for more of the two first years.%
\footnote{
Recall that quarter 0 is equivalent to 9 months (3 quarters) after the birth
of the first child.
}

>>>>

What has not previously been appreciated is that the introduction 
of prenatal sex determination fundamentally changed the relationship 
between son preference and birth spacing.
This is because each sex-selective abortion substantially increases the 
spacing from last birth. 
For every abortion a woman has we should expect the duration from last
birth to increase by approximately a year.
This increase consists of three parts starting from the time of the 
abortion.
First, the uterus needs at least two menstrual cycles to recover 
before conception, because a less than three months space 
between abortion and conception substantially increases the likelihood 
of a spontaneous abortion \citep{zhou00b}.
Second, an average of 6 months to next conception \citep{Wang2003}.
Finally, sex determination tests are generally reliable only from 3 months 
of gestation onwards.
These three parts add up to 12 months.%
\footnote{
The waiting time to conception does vary by woman, but even if it is very 
short, say one  month, the additional space between births would still
be 6 months per abortion.
}

@@

The main downside of this approach is that direct interpretation of the estimated 
coefficients for this model is challenging because of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, 
no birth, rather than simply the hazard of an event.
Second, a positive coefficient does not necessarily imply that an increase in a
variable's value increases the probability of the associated event because the 
probability of another event may increase even more \citep{thomas96}.

@@

Spell duration is measured in months.
The first spell begins at the month of marriage because many 
women report giving birth less than 9 months after they were married.
For women who began living with their husbands at too young an age to conceive, the 
starting point should ideally be first ovulation, when she becomes ``at risk'' for a 
pregnancy, rather than month of marriage.
Unfortunately, information on age of menarche is only available in NFHS-1.
Instead, for women who began living with their husband before age 12, I set the 
the month they turned 12 years of age as staring point for the first spell.

@@

All spells continue until either a child is born or the spell is censored.
Censoring can happen for three reasons:
the survey takes place;
the woman is sterilized;
or the number of births observed becomes too sparse for the method to work.
The timing of censoring because of too few births vary slightly by spell but 
is generally 5 years or longer after the beginning of the spell.

@@

The area of residence is a dummy variable for the household living in
an urban area.%
\footnote{
NFHS uses four categories for area of residence: Large city, small city, town
and countryside.
To reach a sufficient sample size urban areas are merged into one group.
}
The cost of children is higher in urban areas than in rural and access to prenatal
sex determination is easier, and both are expected to be associated with
greater use of sex-selective abortions in urban than in rural areas.
Because of concerns about selective migration I use where the household was living
at the end of each spell.
% %
% \footnote{
% As online Appendix Table D.1 shows there is little difference across
% education levels in migration patterns within each period and patterns across periods
% are also relatively stable.
% }

@@

% The increase in urbanization and education is likely to exert downward pressure
% on fertility and the high censoring rates for the later periods are evidence of this.
% The average number of children born to women by the time they turn 35 illustrate how
% strong the decline in fertility has been.%
% \footnote{
% Figure \ref{fig:fertility} shows the full results.
% }
% Women born in the early 1940s had on average close to 5 children when 
% they reached 35, but women born in the early 1970s had only just over three children.
% The low number of children is especially remarkable because it combines
% all education levels and all areas of residence.
% % Hence, fertility in cities is likely substantially lower.

@@

The main question of interest here is how the duration between one
birth and the next is related to son preference.
The hypothese examined here is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
used.
We would still expect parents to \emph{begin} trying to conceive earlier
after the birth of a girl than after a boy, but the abortions will automatically
increase the length of the space, and may even lead to longer spacing after the
birth of girls than after boys.

To provide a first look at how the hypothese holds up, Tables 
\ref{tab:median_sex_ratio_low},
\ref{tab:median_sex_ratio_med}, and
\ref{tab:median_sex_ratio_high}
show predicted median duration and sex ratios for the three different
education levels by spell, sex composition, and periods.
These predictions are based on the estimation results employing the method 
detailed in Section \ref{sec:strategy} and are calculated using the means of 
the continuous explanatory variables and the majority category for categorical 
explanatory variables.
The median duration is the number of months it is predicted to take for
half of the women who are predicted to have a birth by the end of spell length
covered.
For example, if 80\% of women with a given set of characteristics are predicted to
have a birth by the end of the end of spell, the median duration is the number of
months it would take for 40\% of women to have a birth.
The sex ratio is the percent of births that result in a son, based on all the 
births predicted to happen over the spell.

@@

Figure \ref{fig:results_spell2_high_urban} 
shows the predicted probabilities of having a boy and standard survival curves
for the three periods for urban women with 8 or more years of education.%
\footnote{
The corresponding graphs for rural women and for women with no education are
shown in Figures \ref{fig:results_spell2_high_rural}, \ref{fig:results_spell2_low_urban},
and \ref{fig:results_spell2_low_rural}
}
Two main results stand out.
First, women whose first child was a boy do not appear to use sex selection during
the second spell.
The predicted percent boys is consistently close to the natural level in all three
panels (\ref{fig:results_spell2_high_urban}(d)-(f)).
Second, as prenatal sex determination became available there is more and more
evidence of sex selection affecting the sex ratio for women with a girl as their
first child.
In panel (a), which covers the period before sex selection became generally
available, there is no significant evidence of a sex ratio biased towards males.
Panel (b) shows a more male biased sex ratio but only a few months show
a statistically significant difference from the natural rate and the shallow 
slope of the survival curve indicates that relatively few births took place
during these months.
Panel (c), however, shows clear evidence of the effect of access to prenatal
sex determination with a male biased sex ratio for more of the two first years.%
\footnote{
Recall that month 0 is equivalent to 9 months after the birth
of the first child.
}
